On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-11 paper/linear3d.tex
regular update

%On the other hand, if a time-independent heat source is invoked, then
%we may choose $\alpha, \beta$ and $\theta$ separately.  
\section{Linear problem}
We consider infinitestimal axisymmetric Eulerian perturbations in the
form \begin{align}
  \delta \rho= \widetilde{\delta \rho}(z)\exp{\left(\ii kx +
    s t\right)},  
\end{align}
and similarly for other variables. Here, $k$ 
is a real horizontal wavenumber and $s$ is a (generally) complex
growth rate. 
%The gravitational potential perturbation $\delta \Phi$ depends on $z$
%but only its midplane value $\delta \Phi(z=0)$ is
%used in the dynamical equations, which has the same form as above. 
For clarity, hereafter we drop the tilde.  
%and note that equilibrium
%quantities without the $\delta$ prefix.

The linearized continuity, momentum and energy equations are
\begin{align}
   &s \delta \rho = - \ii k \rho\delta v_x -
  \left(\rho\delta v_z\right)^\prime \label{lin_mass} \\ 
   &s  \delta v_x = - \ii k\frac{\delta P}{\rho} - \ii k \delta 
  \Phi + 2\Omega \delta v_y + \delta F_x,\\
   &s \delta v_y = (q-2)\Omega\delta v_x + \delta F_y,\\
%   &s \delta v_z = - \left(\frac{\delta P}{\rho}\right)^\prime -
%  \delta\Phi^\prime +
%  \left(\ln{\rho}\right)^\prime\left(c_s^2\frac{\delta\rho}{\rho} -
%  \frac{\delta P}{\rho}\right) + \delta F_z,\\
  &s \delta v_z =
  \left(\ln{P}\right)^\prime c_s^2\frac{\delta\rho}{\rho} -
  \left(\frac{\delta P}{\rho}\right)^\prime -
  \left(\ln{\rho}\right)^\prime\frac{\delta P}{\rho} + \delta F_z, \\ 
  &s \frac{\delta P}{\rho} = 
%- \ii \gamma k \frac{c_s^2}{\Gamma}\delta v_x 
 - \ii k \gamma c_s^2\delta v_x
%- c_s^2\left[
%  \left(\ln{\rho}\right)^\prime\delta v_z + 
%  \frac{\gamma}{\Gamma}\delta v_z^\prime\right] \notag\\
- c_s^2\left[
 \left(\ln{P}\right)^\prime\delta v_z + 
  \gamma\delta v_z^\prime\right] \notag\\
   & \phantom{s \frac{\delta P}{\rho}=}
  + (\gamma-1)\frac{\delta\mathcal{H}_\mathrm{visc}}{\rho} 
  -(\gamma-1)\frac{\delta\Lambda}{\rho}
%  - \frac{1}{t_c}\left(\frac{\delta P}{\rho} -
%  \frac{\theta}{\Gamma}c_s^2\frac{\delta\rho}{\rho}\right) 
  \label{lin_energy},\\
  &\delta\Phi^{\prime\prime} - k^2\delta\Phi =
  \frac{\Omega^2}{Q_\mathrm{3D}}\left(\frac{\delta\rho}{\rho_0}\right)\label{lin_gravity},  
\end{align}
with
\begin{align}
  \delta F_x =& \nu \left[\delta v_x^{\prime\prime} + 
    \left(\ln{\rho\nu}\right)^\prime \delta v_x^\prime - 
     \frac{4}{3}k^2\delta v_x \right] - \nu_bk^2\delta v_x \notag\\
   &+ \ii \nu k \left[\frac{1}{3}\delta v_z^\prime+ 
    \left(\ln{\rho\nu}\right)^\prime\delta v_z\right] + \ii k \nu_b
  \delta v_z^\prime,\\
  \delta F_y = & \nu  \left[\delta v_y^{\prime\prime} + 
    \left(\ln{\rho\nu}\right)^\prime \delta v_y^\prime - 
     k^2\delta v_y \right] - \ii\nu kq\Omega\delta\ln{\rho\nu},\\
  \delta F_z =& \nu\left[\frac{4}{3}\delta v_z^{\prime\prime} + 
    \frac{4}{3}\left(\ln{\rho\nu}\right)^\prime \delta v_z^\prime - 
    k^2\delta v_z \right] \notag\\
  & +\nu_b\left[\delta v_z^{\prime\prime} + \left(\ln{\rho\nu_b}\right)^\prime\delta v_z^\prime\right]\notag\\
  & + \ii \nu k  \left[\frac{1}{3}\delta v_x^\prime -  
    \frac{2}{3}\left(\ln{\rho\nu}\right)^\prime\delta v_x\right]\notag\\
  &+\ii \nu_b k \left[\delta v_x^\prime + \left(\ln{\rho\nu_b}\right)^\prime\delta v_x\right],\\
%  &\delta f_x = -\frac{4}{3} \nu k^2\Sigma \delta v_x,\\
%  &\delta f_y = -\ii q k \Omega \delta(\nu\Sigma) - \nu k^2
%  \Sigma\delta v_y,\\
%  &\delta Q_\mathrm{visc} = (q\Omega)^2\delta(\nu\Sigma) - 2 \ii \nu 
%  k q\Omega\Sigma\delta v_y,\\
%  &\delta(\nu\Sigma) =
%  \frac{\alpha}{\Omega}\left[c_s^2(1+\mu)\delta\Sigma + \lambda\delta P\right].
%  &\delta f_y = - \ii \nu q k \Omega \delta \Sigma - \nu k^2\Sigma\delta
%  v_y,\\
%  &\delta Q_\mathrm{visc} = \nu
%  \left(q\Omega\right)^2\delta\Sigma - 2\ii\nu q k \Omega\Sigma\delta
%  v_y. 
  \frac{\delta\mathcal{H}_\mathrm{visc}}{\rho}=& \nu (q\Omega)^2
  \delta\ln{\rho\nu} - 2\ii\nu k q \Omega \delta v_y,\\
  \delta\ln{\rho\nu} =& (1+\mu)\frac{\delta\rho}{\rho} +
  \frac{\lambda}{c_s^2}\frac{\delta P}{\rho}, 
\end{align}\label{linear_beta_cool}
where $^\prime$ denotes $d/dz$. The linearized cooling functions
$\delta\Lambda$ are given below. 
Eq. \ref{lin_mass}---\ref{lin_gravity}, supplemented with appropriate
boundary conditions, constitutes an eigenvalue problem for the growth
rate $s$.

>>>>

%On the other hand, if a time-independent heat source is invoked, then
%we may choose $\alpha, \beta$ and $\theta$ separately.  
\section{Linear problem}\label{linear}
We consider infinitestimal axisymmetric Eulerian perturbations in the
form \begin{align}
  \delta \rho= \widetilde{\delta \rho}(z)\exp{\left(\ii kx +
    s t\right)},  
\end{align}
and similarly for other variables. Here, $k$ 
is a real horizontal wavenumber and $s$ is a (generally) complex
growth rate. 
%The gravitational potential perturbation $\delta \Phi$ depends on $z$
%but only its midplane value $\delta \Phi(z=0)$ is
%used in the dynamical equations, which has the same form as above. 
For clarity, hereafter we drop the tilde.  
%and note that equilibrium
%quantities without the $\delta$ prefix.

The linearized continuity, momentum and energy equations are
\begin{align}
   &s \delta \rho = - \ii k \rho\delta v_x -
  \left(\rho\delta v_z\right)^\prime \label{lin_mass} \\ 
   &s  \delta v_x = - \ii k\frac{\delta P}{\rho} - \ii k \delta 
  \Phi + 2\Omega \delta v_y + \delta F_x,\\
   &s \delta v_y = (q-2)\Omega\delta v_x + \delta F_y,\\
%   &s \delta v_z = - \left(\frac{\delta P}{\rho}\right)^\prime -
%  \delta\Phi^\prime +
%  \left(\ln{\rho}\right)^\prime\left(c_s^2\frac{\delta\rho}{\rho} -
%  \frac{\delta P}{\rho}\right) + \delta F_z,\\
  &s \delta v_z =
  \left(\ln{P}\right)^\prime c_s^2\frac{\delta\rho}{\rho} -
  \left(\frac{\delta P}{\rho}\right)^\prime -
  \left(\ln{\rho}\right)^\prime\frac{\delta P}{\rho} + \delta F_z, \\ 
  &s \frac{\delta P}{\rho} = 
%- \ii \gamma k \frac{c_s^2}{\Gamma}\delta v_x 
 - \ii k \gamma c_s^2\delta v_x
%- c_s^2\left[
%  \left(\ln{\rho}\right)^\prime\delta v_z + 
%  \frac{\gamma}{\Gamma}\delta v_z^\prime\right] \notag\\
- c_s^2\left[
 \left(\ln{P}\right)^\prime\delta v_z + 
  \gamma\delta v_z^\prime\right] \notag\\
   & \phantom{s \frac{\delta P}{\rho}=}
  + (\gamma-1)\frac{\delta\mathcal{H}_\mathrm{visc}}{\rho} 
  -(\gamma-1)\frac{\delta\Lambda}{\rho}
%  - \frac{1}{t_c}\left(\frac{\delta P}{\rho} -
%  \frac{\theta}{\Gamma}c_s^2\frac{\delta\rho}{\rho}\right) 
  \label{lin_energy},\\
  &\delta\Phi^{\prime\prime} - k^2\delta\Phi =
  \frac{\Omega^2}{Q_\mathrm{3D}}\left(\frac{\delta\rho}{\rho_0}\right)\label{lin_gravity},  
\end{align}
with
\begin{align}
  \delta F_x =& \nu \left[\delta v_x^{\prime\prime} + 
    \left(\ln{\rho\nu}\right)^\prime \delta v_x^\prime - 
     \frac{4}{3}k^2\delta v_x \right] - \nu_bk^2\delta v_x \notag\\
   &+ \ii \nu k \left[\frac{1}{3}\delta v_z^\prime+ 
    \left(\ln{\rho\nu}\right)^\prime\delta v_z\right] + \ii k \nu_b
  \delta v_z^\prime,\\
  \delta F_y = & \nu  \left[\delta v_y^{\prime\prime} + 
    \left(\ln{\rho\nu}\right)^\prime \delta v_y^\prime - 
     k^2\delta v_y \right] - \ii\nu kq\Omega\delta\ln{\rho\nu},\\
  \delta F_z =& \nu\left[\frac{4}{3}\delta v_z^{\prime\prime} + 
    \frac{4}{3}\left(\ln{\rho\nu}\right)^\prime \delta v_z^\prime - 
    k^2\delta v_z \right] \notag\\
  & +\nu_b\left[\delta v_z^{\prime\prime} + \left(\ln{\rho\nu_b}\right)^\prime\delta v_z^\prime\right]\notag\\
  & + \ii \nu k  \left[\frac{1}{3}\delta v_x^\prime -  
    \frac{2}{3}\left(\ln{\rho\nu}\right)^\prime\delta v_x\right]\notag\\
  &+\ii \nu_b k \left[\delta v_x^\prime + \left(\ln{\rho\nu_b}\right)^\prime\delta v_x\right],\\
%  &\delta f_x = -\frac{4}{3} \nu k^2\Sigma \delta v_x,\\
%  &\delta f_y = -\ii q k \Omega \delta(\nu\Sigma) - \nu k^2
%  \Sigma\delta v_y,\\
%  &\delta Q_\mathrm{visc} = (q\Omega)^2\delta(\nu\Sigma) - 2 \ii \nu 
%  k q\Omega\Sigma\delta v_y,\\
%  &\delta(\nu\Sigma) =
%  \frac{\alpha}{\Omega}\left[c_s^2(1+\mu)\delta\Sigma + \lambda\delta P\right].
%  &\delta f_y = - \ii \nu q k \Omega \delta \Sigma - \nu k^2\Sigma\delta
%  v_y,\\
%  &\delta Q_\mathrm{visc} = \nu
%  \left(q\Omega\right)^2\delta\Sigma - 2\ii\nu q k \Omega\Sigma\delta
%  v_y. 
  \frac{\delta\mathcal{H}_\mathrm{visc}}{\rho}=& \nu (q\Omega)^2
  \delta\ln{\rho\nu} - 2\ii\nu k q \Omega \delta v_y,\\
  \delta\ln{\rho\nu} =& (1+\mu)\frac{\delta\rho}{\rho} +
  \frac{\lambda}{c_s^2}\frac{\delta P}{\rho}, \label{linear_beta_cool}
\end{align}
where $^\prime$ denotes $d/dz$, and the 3D self-gravity parameter is
\begin{align}
  Q_\mathrm{3D} \equiv \frac{\Omega^2}{4\pi G \rho_0}. 
\end{align}
The linearized cooling functions $\delta\Lambda$ are given below. 
Eq. \ref{lin_mass}---\ref{lin_gravity}, supplemented with appropriate
boundary conditions, constitutes an eigenvalue problem for the growth
rate $s$.

### 2016-01-11 paper/model3d.tex
regular update

\section{Basic equations}
We consider a three-dimensional (3D), self-gravitating,
viscous disk with heating and cooling. We use the shearing box framework to
study a small patch of the disk \citep{goldreich65}. The local frame
co-rotates with a fiducial point in the unperturbed disk at angular
frequency $\Omega$. The Cartesian co-ordinates $(x,y,z)$ corresponds
to the radial, azimuthal and verical directions in the global disk. 
The fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) = 0, \\
  & \frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi - 2\Omega\hat{\bm{z}}\times\bm{v}\notag\\ & 
  \phantom{\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =}
  +2\Omega^2qx\hat{\bm{x}} -\Omega_z^2z\hat{\bm{z}} 
   + \frac{1}{\rho}\nabla\cdot\bm{T},\label{momentum_eq}\\
  & \frac{\p E}{\p t} + \nabla\cdot\left(E\bm{v}\right) = - P
  \nabla\cdot\bm{v} + \mathcal{H}_\mathrm{visc} +
  \mathcal{H}_\mathrm{ext} - \Lambda,\label{energy_eq}
\end{align} 
where $\rho$ is the density field and $\bm{v} = (v_x, v_y, v_z)$ is
the velocity field.% $q$ is the shear parameter and $\Omega_z$ is the
%local vertical oscillation frequency. 
We assume an ideal has so that the pressure $P$ and thermal energy
density $E$ are related by  
\begin{align}
  P = (\gamma-1)E = \mathcal{R}\rho T, 
\end{align}
where $\gamma$ is the adiabatic index, $\mathcal{R}$ is the gas
constant, and $T$ is the temperature. We use $\gamma=1.4$ unless
otherwise stated. The gas 
gravitatational potential $\Phi$ is given via the Poisson equation,  
\begin{align}
 \nabla^2\Phi = 4 \pi G \rho, 
\end{align}
where $G$ is the gravitational constant.

@@

In the energy equation (Eq. \ref{energy_eq}) the source terms 
$\mathcal{H}_\mathrm{visc}$, $\mathcal{H}_\mathrm{ext}$ and $\Lambda$
represent viscous heating, time-independent external heating and
thermal losses, respectively, described below. We normally set
$\mathcal{H}_\mathrm{ext}=0$ unless otherwise stated.

@@

We adopt the viscosity law  
\begin{align}
  \nu = \alpha
  \frac{c_{s0}^2}{\Omega}\left(\frac{\rho}{\rho_\mathrm{eq}}\right)^\mu\left(\frac{P}{P_\mathrm{eq}}\right)^\lambda,            
\end{align}
where subscript `eq' denotes the equilibrium state and  
%$c_{s0}$ is a modified sound-speed defined below. 
$c_{s0}^2\equiv P_\mathrm{eq}(0)/\rho_\mathrm{eq}(0)$. 
The dimensionless coefficient
$\alpha=\alpha(\rho_\mathrm{eq},P_\mathrm{eq})$ characterizes the
magnitude of the shear viscosity and is time-independent.  
The indices $\mu,\,\lambda$ are chosen to model how 
the viscosity behaves in the perturbed state. Given the uncertain
properties of gravito-turbulence, we will treat $\mu,\lambda$ as free
parameters. We adopt the same prescription for the bulk viscosity but
with $\alpha\to\alpha_b$.

Our numerical calculations mostly use $\mu=-1,\lambda=0$ so
that $\rho\nu$ remains constant, as considered
in previous gravitational instabilities with 
viscosity \citep{hunter83,gammie96}. 
This model conveniently eliminates  
viscous over-stability \citep{schmit95,latter06}, which is unrelated
to self-gravity, that would otherwise contaminate our results.

While the choice $\mu=-1,\lambda=0$ may seem artifical, we emphasize
this is done in order to bring out the relevant physical process ---
viscosity-assisted gravitational instability. Interestingly, 
\cite{laughlin96b} had already suggested a $1/\rho$ dependence for the 
shear viscosity when modeling the evoluting of self-gravitating disks
as a viscous process.  
{\bf note paardekooper 2012: alpha goes down as Sigma goes up}

@@

\subsection{Parameters}
{\bf temporary, maybe move to intro}
For the 3D problem it is convenient to parameterize the strength of
self-gravity via 
\begin{align}
  Q_\mathrm{3D} \equiv \frac{\Omega^2}{4\pi G \rho_0}. 
\end{align}
It is also useful to define the
two-dimensional (2D) Toomre parameter 
\begin{align}
  Q = \frac{c_{s0}\Omega}{\pi G \Sigma}, 
\end{align}
where $\Sigma = \int_{-\infty}^\infty\rho dz$ is the surface density.

>>>>

\section{Basic equations}\label{basic}
We consider a three-dimensional (3D), self-gravitating,
viscous disk with heating and cooling. We use the shearing box framework to
study a small patch of the disk \citep{goldreich65}. The local frame
co-rotates with a fiducial point in the unperturbed disk at angular
frequency $\Omega$. The Cartesian co-ordinates $(x,y,z)$ corresponds
to the radial, azimuthal and verical directions in the global disk. 
The fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) = 0, \\
  & \frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi - 2\Omega\hat{\bm{z}}\times\bm{v}\notag\\ & 
  \phantom{\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =}
  +2\Omega^2qx\hat{\bm{x}} -\Omega_z^2z\hat{\bm{z}} 
   + \frac{1}{\rho}\nabla\cdot\bm{T},\label{momentum_eq}\\
  & \frac{\p E}{\p t} + \nabla\cdot\left(E\bm{v}\right) = - P
  \nabla\cdot\bm{v} + \mathcal{H}_\mathrm{visc} +
  \mathcal{H}_\mathrm{ext} - \Lambda,\label{energy_eq}
\end{align} 
where $\rho$ is the density field and $\bm{v} = (v_x, v_y, v_z)$ is
the velocity field.% $q$ is the shear parameter and $\Omega_z$ is the
%local vertical oscillation frequency. 
We assume an ideal gas so that the pressure $P$ and thermal energy
density $E$ are related by  
\begin{align}
  P = (\gamma-1)E = \mathcal{R}\rho T, 
\end{align}
where $\gamma$ is the adiabatic index, $\mathcal{R}$ is the gas
constant, and $T$ is the temperature. We use $\gamma=1.4$ unless
otherwise stated. The gas 
gravitatational potential $\Phi$ is given via the Poisson equation,  
\begin{align}
 \nabla^2\Phi = 4 \pi G \rho, 
\end{align}
where $G$ is the gravitational constant.

@@

In the energy equation (Eq. \ref{energy_eq}) the source terms 
$\mathcal{H}_\mathrm{visc}$, $\mathcal{H}_\mathrm{ext}$ and $\Lambda$
represent viscous heating, any time-independent heat
source/sinks, and time-dependent thermal losses, respectively. These
are described below. We normally set
$\mathcal{H}_\mathrm{ext}=0$ unless otherwise stated.

@@

We adopt the viscosity law  
\begin{align}
  \nu = \alpha
  \frac{c_{s0}^2}{\Omega}\left(\frac{\rho}{\rho_\mathrm{eq}}\right)^\mu\left(\frac{P}{P_\mathrm{eq}}\right)^\lambda,            
\end{align}
where subscript `eq' denotes the equilibrium state and  
%$c_{s0}$ is a modified sound-speed defined below. 
$c_{s0}^2\equiv P_\mathrm{eq}(0)/\rho_\mathrm{eq}(0)$. 
The dimensionless coefficient
$\alpha=\alpha(\rho_\mathrm{eq},P_\mathrm{eq})$ characterizes the
magnitude of the shear viscosity \emph{in steady state} 
and is time-independent. The indices $\mu,\,\lambda$ are free
parameters chosen to model how the viscosity behaves in the perturbed
state.  
%Given the uncertain
%properties of gravito-turbulence, we will treat $\mu,\lambda$ as free
%parameters. 
We adopt the same prescription for the bulk viscosity but
with $\alpha\to\alpha_b$.

Our numerical calculations use $\mu=-1,\lambda=0$ so 
that $\rho\nu$ remains constant, as adopted in previous, pioneering
studies of viscous GI 
\citep{lynden-bell74,hunter83,willerding92,gammie96}.  
This choice also conveniently eliminates viscous over-stability
\citep{schmit95,latter06}, which is unrelated to self-gravity, that
would otherwise contaminate our results.

%the alpha(rho_eq, P_eq) function can be derived with some rigor 
%but this assumes steady state. don't know how to model
%time-dependence.

While steady-state viscosity values can be determined analytically
\citep[e.g.][]{rafikov15}, its time-dependent behavior is not
well-explored. \cite{laughlin96b} suggested a $\nu \propto 1/\Sigma$
dependence when modeling the evoluting of self-gravitating disks 
as a viscous process. \cite{paardekooper12} measured
an increase in the effective viscosity when their gravito-turbulent
disk models underwent fragmentation (i.e. as $\Sigma$ increases).  
These motivate the choice $\mu=-1,\lambda=0$, but we emphasize this is
done to bring out the physical process of interest --- viscosity-assisted
gravitational instability.

%{\bf note paardekooper 2012: alpha goes down as Sigma goes up}

@@

%\subsection{Parameters}
%{\bf temporary, maybe move to intro}
%For the 3D problem it is convenient to parameterize the strength of
%self-gravity via 
%\begin{align}
%  Q_\mathrm{3D} \equiv \frac{\Omega^2}{4\pi G \rho_0}. 
%\end{align}
%It is also useful to define the
%two-dimensional (2D) Toomre parameter 
%\begin{align}
%  Q = \frac{c_{s0}\Omega}{\pi G \Sigma}, 
%\end{align}
%where $\Sigma = \int_{-\infty}^\infty\rho dz$ is the surface density.

### 2016-01-11 paper/paper3d.tex
regular update

%\documentclass[12pt]{aastex}
\documentclass[iop, numberedappendix]{emulateapj}
%\documentclass{article}
%\usepackage{draftwatermark}
%\usepackage{paralist}
\shorttitle{Non-ideal GI}
\shortauthors{M-K.\ Lin, K. Kratter}

@@

\begin{abstract}
Low mass, self-gravitating accretion disks are able to sustain a  
quasi-steady, `gravito-turbulent' state provided that thermal 
time-scales are sufficiently long, or equivalently for sufficiently
weak turbulent stresses. Otherwise, numerical simulations show that
disk fragmentation ensues.   
We explore the physical interpretation of this result as the
compromise of gravitational stability by non-ideal effects: 
\begin{inparaenum}[1)] 
\item cooling, which reduces pressure support; and
\item viscosity, which reduces rotational support. 
\end{inparaenum}
We derive a flexible dispersion relation for linear waves in a 
self-gravitating, viscous, non-adiabatic disk with internal and
external heating, and explicit cooling. We 
consider parameterized cooling functions in two and three dimensions,
as well as full radiative diffusion in 3D. 
We apply linear theory to 
global models of protoplanetary disks (PPDs) subject to gravito-turbulence, modeled as an effective
viscosity, and cooling via dust opacity. We find the sole effect of cooling is insufficient to
destabilize a realistic PPD. However, inclusion of viscosity renders
the disk beyond $\sim 60$AU dynamically unstable on radial 
lengthscales a few times the local disk thickness. 
Thus, a steady gravito-turbulent state cannot persist for more and a
few orbits in the outer disk, coincident with the expected region for
fragmentation based on emperical conditions inferred from numerical
simulations. 
%This is understood within our framework as the disk's
%inability to sustain too large of a stress as it undermines
%rotational support against self-gravity. 
%This motivates an interpretation of disk 
%fragmentation as a secondary gravitational instability enabled by 
%the underlying gravito-turbulent viscosity when it exceeds a critical value.         
\end{abstract}
%opacity

@@

This paper is organized as follows.

@@

>>>>

%\documentclass[12pt]{aastex}
\documentclass[iop, numberedappendix]{emulateapj}
%\documentclass{article}
%\usepackage{draftwatermark}
%\usepackage{paralist}
\shorttitle{Non-ideal Gravitational instability}
\shortauthors{M-K.\ Lin \& K. Kratter}

@@

\begin{abstract}
Low mass, self-gravitating accretion disks can remain in a  
quasi-steady, `gravito-turbulent' state provided that thermal 
time-scales are sufficiently long, or equivalently for sufficiently
weak turbulent stresses. Otherwise, numerical simulations show that
disk fragmentation ensues. 
We quantify the interpretation of this result as the
compromise of gravitational stability by non-ideal physics: 
\begin{inparaenum}[1)] 
\item cooling, which reduces pressure support; and
\item viscosity, which reduces rotational support. 
\end{inparaenum}
To this end, we analyse the gravitational stability of accretion disks
including these effects explicitly.   
We derive a versatile dispersion relation for linear waves in a 
self-gravitating, viscous, non-adiabatic disk with internal and
external heating, and explicit cooling. We 
consider parameterized cooling functions in two and three dimensions,
as well as full radiative diffusion in 3D. 
We apply linear theory to 
global models of protoplanetary disks (PPDs) subject to gravito-turbulence, modeled as an effective
viscosity, and cooling via dust opacity. We find the sole effect of cooling is insufficient to
destabilize a realistic PPD. However, inclusion of viscosity renders
the disk beyond $\sim 60$AU dynamically unstable on radial 
lengthscales a few times the local disk thickness. This is coincident 
with the numerically-established empirical condition for disk fragmentation based on a
maximum sustainable stress. We 
thus suggest that disk fragmentation can be physically attributed,
at least partially, to the removal of rotational stabilization against
self-gravity by turbulence.   
\end{abstract}
%alternative abstract

@@

The basic equations are listed in \S\ref{basic}, along with
descriptions of the disk 
equilibra and the two cooling models 
considered --- parametrized `beta' cooling and radiative diffusion. 
The linear stability problem is defined in
\S\ref{linear}. We present results with beta cooling in
\S\ref{result_2d} and \S\ref{3ddisk} for 2D and 3D disks, respectively. In
\S\ref{2dppd} we apply linear theory to PPDs with realistic cooling
models, including radiative diffusion in 3D. We find regions in a
PPD that cannot sustain steady gravito-turbulence to coincide with
that expected for disk fragmentation, based on previous empirical
criteria. We summarize our results in \S\ref{summary}, including a
discussion of how our model may aid the physical understanding of  
fragmentation in realistic PPDs. 
% as the inability to sustain gravito-turbulence
%because the associated stress removes rotational support

@@

%\acknowlegement
%S. Stahler

### 2016-01-11 paper/summary.tex
regular update

\section{Summary and discussion}
In this paper, we study the implications on the stability of
turbulent, cooling accretion disks if said turbulence is modeled as an
effective, Navier-Stokes viscosity. This viscosity provides a 
background heating and may also be active in the perturbed 
state. We pay particular attention to 
protoplanetary disks subject to `gravito-turbulence'. 
Our framework provide an alternative interpretation of existing
results on PPD fragmentation established from direct numerical
simulations. 
{\bf cooling removes pressure support, viscosity removes rotational support}
{\bf our conclusions do require input of emperical resuls from numerical
sims}

>>>>

\section{Summary and discussion}\label{summary}
In this paper, we study the implications on the stability of
turbulent, cooling accretion disks if said turbulence is modeled as an
effective, Navier-Stokes viscosity. This viscosity provides a 
background heating and may also be active in the perturbed 
state. We pay particular attention to 
protoplanetary disks subject to `gravito-turbulence'. 
Our framework provide an alternative interpretation of existing
results on PPD fragmentation established from direct numerical
simulations. 
{\bf cooling removes pressure support, viscosity removes rotational support}
{\bf our conclusions do require input of emperical resuls from numerical
sims}
